Gaussian Processes Autoregressive Models for Forecasting the Disturbance Storm Time Index

### 2016-05-23 swsc.tex
Updates from ShareLaTeX

Here $\matbf{f}$ is an $N\times 1$ vector consisting of the values $f(\mathbf{x}_i), i = q \cdots N$. In equation \ref{eq:normal} $\mathbf{f}|\mathbf{x}_1, \cdots, \mathbf{x}_N$ denotes that we are modeling the conditional distribution of $\mathbf{f}$ with respect to the input data (i.e. $\mathbf{X}$). $\mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)$ represents a multivariate Gaussian distribution with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Lambda}$.

@@

With the multivariate normal distribution defined in equation \ref{eq:dist}, probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly as equation \ref{eq:posterior} (see \citet{Rasmussen:2005:GPM:1162254}).

@@

@@

The central design issue now becomes choice of the function $K(\mathbf{x}, \mathbf{x}')$. The same constraints that apply to $\mathbf{\Lambda}$ also apply to the function $K(.,.)$. In machine learning, these symmetric positive definite functions of two variables are known as \emph{kernels}. Kernel based methods are applied extensively in data analysis i.e. regression, clustering, classification, density estimation (see \citet{Scholkopf:2001:LKS:559923}, \citet{hofmann2008}).

@@

The quantities $l$ in the RBF, $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The method of assigning values to the \emph{hyper-parameter} is crucial in the model building process (see section \ref{sec:hyp} below). In this study, we construct Gaussian Process regression models with polynomial kernels for \emph{one step ahead} prediction of the $Dst$ index.

\begin{comment}
\subsubsection{Radial Basis Function (RBF)}

The RBF kernel is the most commonly used kernel in machine learning, in its core it encodes the idea of \emph{smoothness} because functions drawn from its Hilbert space are infinitely differentiable. The RBF kernel is also an example of a \emph{stationary} kernel i.e. its value depends only on $||\mathbf{x} - \mathbf{y}||$. The quantity $l$ is known as the characteristic \emph{length scale} of this kernel, it is a free parameter (also known as hyper-parameter) which must be determined by the modeller, refer to section \ref{sec:hyp} on how suitable values of kernel hyper-parameters.

\begin{equation}
    K(\mathbf{x}, \mathbf{y}) = \frac{1}{2} exp(-||\mathbf{x} - \mathbf{y}||^2/l^2)
\end{equation}

\subsubsection{Polynomial}

The polynomial kernel is also a common kernel function employed in research, it has two hyper-parameters the degree $d$ and an intercept $b$. The polynomial kernel formulates the function approximation in terms of polynomial expansions of the input dimensions $\mathbf{x}$, more specifically it considers all monomial combination of the input dimensions up to degree $d$. When $d = 1$ it corresponds to a Bayesian linear regression model.

\begin{equation}
    K(\mathbf{x}, \mathbf{y}) = (\mathbf{x}^\intercal \mathbf{y} + b)^d 
\end{equation}

\end{comment}

@@

The simplest way to perform \emph{model selection} is called \emph{grid search} which entails evaluating the above criteria on a fixed grid which is constructed by making a list of values for each hyper-parameter and taking the combinations of all such values. There are more sophisticated methods for \emph{model tuning} such as \emph{coupled simulated annealing} (\citet{Xavier-De-Souza2010}) and \emph{simplex search} (\citet{Nelder1965}) but they are computationally more expensive.

@@

The simplest auto-regressive models for \emph{OSA} prediction of $Dst$ are those that use only the history of $Dst$ to construct input features for model training. The input features $\mathbf{x}_{t-1}$ at each time step is the history of $Dst(t)$ until a time lag of $p$ steps.

@@

Auto-regressive models can be improved by including exogenous quantities that can help in improving predictions. It is well known that the solar wind and interplanetary magnetic field are important factors that influence the Earth's geomagnetic state.

Hourly solar wind velocity $V(t)$ and interplanetary magnetic field $B_z(t)$ recorded at the ACE satellite are used as the exogenous inputs of the model, along with the time history of $Dst$. In this model we choose distinct time lags $p$ and $p_{ex}$ for auto-regressive and exogenous variables respectively.

@@

Before performance bench marks for \emph{OSA} $Dst$ prediction are run on the storm events in \cite{Ji2012}, training and model selection of \emph{GP-AR} and \emph{GP-ARX} models on independent data sets must be performed. For this purpose we choose segments 00:00 1 January 2008 - 10:00 11 January 2008 for training and 00:00 15 November 2014 - 23:00 1 December 2014 for model selection.

@@

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_RMSE.png}
      \caption{Averaged RMSE performance of GP-AR, GP-ARX and the Persistence model versus results in \citet{Ji2012}.}
         \label{fig:rmse}
   \end{figure}

@@

\citet{Ji2012} compare a six OSA $Dst$ models in their paper by compiling a list of 63 geomagnetic storm events of varying strengths which occurred from 1998-2006. This serves as an important stepping stone for a systematic comparison of existing and new prediction techniques because performance metrics are averaged over a large number of storm events that occurred over a 8 year period. They compare the averaged performance of these models on four key performance metrics.

\begin{enumerate}
    \item $RMSE$: The root mean square error
    \item $CC$: Correlation coefficient between the predicted and actual value of $Dst$
    \item $\Delta Dst_{min}$: The error in prediction of the peak value of $Dst$ for a storm event.
    \item $|\Delta t_{peak}|$: The error in predicting the timing of a storm peak, also called \emph{timing error}. 
\end{enumerate}

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_RelProb.png}
   \caption{Comparison of empirical distribution of relative errors for GP-AR, GP-ARX, NM and TL on the test set of \citet{Ji2012}.}
   \label{fig:relprob}
\end{figure}

Apart from the metrics above, we also generate the empirical distribution of absolute relative errors $|\frac{Dst - \hat{Dst}}{Dst}|$ for the \emph{GP-AR}, \emph{GP-ARX}, \emph{NM} and \emph{TL} models. These empirical cumulative distributions enable us to upper bound the percentage error in $Dst$ prediction and attach probability estimates for said bounds.

\subsection{Models Compared}

>>>>

Here $\matbf{f}$ is an $N\times 1$ vector consisting of the values $f(\mathbf{x}_i), i = q \cdots N$. In equation \ref{eq:normal} $\mathbf{f}|\mathbf{x}_1, \cdots, \mathbf{x}_N$ denotes the conditional distribution of $\mathbf{f}$ with respect to the input data (i.e. $\mathbf{X}$) and $\mathcal{N}\left( \mathbf{\mu}, \mathbf{\Lambda} \right)$ represents a multivariate Gaussian distribution with mean vector $\mathbf{\mu}$ and covariance matrix $\mathbf{\Lambda}$. The probability density function of this distribution $p( \mathbf{f} \ | \ \mathbf{x}_1, \cdots, \mathbf{x}_N)$ is therefore given by equation \ref{eq:sto}.

@@

With the multivariate normal distribution defined in equation \ref{eq:dist}, probabilistic predictions $f_*$ can be generated by constructing the conditional distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. Since the original distribution of $\left( \begin{array}{c} \mathbf{y} \\ \mathbf{f_*} \end{array} \right) | \ \ \mathbf{X}, \mathbf{X}_*$ is a multivariate Gaussian, conditioning on a subset of elements $\mathbf{y}$ yields another Gaussian distribution whose mean and covariance can be calculated exactly in equation \ref{eq:posterior} (see \citet{Rasmussen:2005:GPM:1162254}).

@@

The practical implementation of \emph{Gaussian Process} models requires the inversion of the training data kernel matrix $[\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1}$ to calculate the parameters of the predictive distribution $\mathbf{f_*}|\mathbf{X},\mathbf{y},\mathbf{X_*}$. The computational complexity of this inference is dominated by the linear problem $\mathbf{K}^T_{*} [\mathbf{K} + \sigma^{2} \mathbf{I}]^{-1} \mathbf{y}$, this can be done via Cholesky decomposition which has time complexity $O(N^3)$.

@@

The central design issue in applying \emph{Gaussian Process} models is choice of the function $K(\mathbf{x}, \mathbf{x}')$. The same constraints that apply to $\mathbf{\Lambda}$ also apply to the function $K(.,.)$. In machine learning, these symmetric positive definite functions of two variables are known as \emph{kernels}. Kernel based methods are applied extensively in data analysis i.e. regression, clustering, classification, density estimation (see \citet{Scholkopf:2001:LKS:559923}, \citet{hofmann2008}).

@@

The quantities $l$ in the RBF, $b$ and $d$ in the polynomial kernel are known as \emph{hyper-parameters}. Hyper-parameters give flexibility to particular kernel structure, for example $d = 1, 2, 3, \cdots$ in the polynomial kernel represents linear, quadratic, cubic and higher order polynomials respectively. The method of assigning values to the \emph{hyper-parameter} is crucial in the model building process (see section \ref{sec:hyp} below). In this study, we construct Gaussian Process regression models with linear polynomial kernels for \emph{one step ahead} prediction of the $Dst$ index.

@@

The simplest way to perform \emph{model selection} is called \emph{grid search} which entails evaluating the above criteria on a fixed grid which is constructed by the tensorial product of grids constructed on each hyper-parameter. There are more sophisticated methods for \emph{model tuning} such as \emph{coupled simulated annealing} (\citet{Xavier-De-Souza2010}) and \emph{simplex search} (\citet{Nelder1965}) but they are computationally more expensive.

@@

The simplest auto-regressive models for \emph{OSA} prediction of $Dst$ are those that use only the history of $Dst$ to construct input features for model training. The input features $\mathbf{x}_{t-1}$ at each time step are the history of $Dst(t)$ until a time lag of $p$ hours.

@@

Auto-regressive models can be improved by including exogenous quantities that can help in improving predictions. It is well known that the solar wind and interplanetary magnetic field are important factors that influence the Earth's geomagnetic state (\citet{JGRA:JGRA11775}, \citet{Russell1974}).

Hourly solar wind velocity $V(t)$ and interplanetary magnetic field $B_z(t)$ recorded by the ACE satellite are used as the exogenous inputs of the model, along with the time history of $Dst$. In this model we choose distinct time lags $p$ and $p_{ex}$ for auto-regressive and exogenous variables respectively. In addition we explicitly include the product $V B_z$ which is a proxy for the electric field, as an input.

@@

Before running performance bench marks for \emph{OSA} $Dst$ prediction on the storm events in \cite{Ji2012}, training and model selection of \emph{GP-AR} and \emph{GP-ARX} models on independent data sets must be performed. For this purpose we choose segments 00:00 1 January 2008 - 10:00 11 January 2008 for training and 00:00 15 November 2014 - 23:00 1 December 2014 for model selection.

@@

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_RMSE.png}
      \caption{Averaged RMSE performance of \emph{GP-AR}, \emph{GP-ARX} and the \emph{Persistence} model versus results in \citet{Ji2012}.}
         \label{fig:rmse}
   \end{figure}

@@

\citet{Ji2012} compare six OSA $Dst$ models in their paper by compiling a list of 63 geomagnetic storm events of varying strengths which occurred from 1998-2006. This serves as an important stepping stone for a systematic comparison of existing and new prediction techniques because performance metrics are averaged over a large number of storm events that occurred over a 8 year period. They compare the averaged performance of these models on four key performance metrics.

\begin{enumerate}
    \item The root mean square error.
    \begin{equation}
        RMSE = \sqrt{\sum_{i=1}^{n} (y - \hat{y})^2 / n}
    \end{equation}
    \item Correlation coefficient between the predicted and actual value of $Dst$.
    \begin{equation}
        CC = Cov(y \hat{y})/\sqrt{Var(y) Var(\hat{y})}
    \end{equation}
    \item The error in prediction of the peak negative value of $Dst$ for a storm event.
    \begin{equation}
        \Delta Dst_{min} = Dst(t^*) - \hat{D}st(t^*), \ t^* = argmin(Dst(t))
    \end{equation}
    \item The error in predicting the timing of a storm peak, also called \emph{timing error} $|\Delta t_{peak}|$. 
\end{enumerate}

\begin{figure}
   \centering
   \includegraphics[width=\textwidth]{Compare_RelProb.png}
   \caption{Comparison of empirical distribution of relative errors for \emph{GP-AR}, \emph{GP-ARX}, \emph{NM} and \emph{TL} on the test set of \citet{Ji2012}.}
   \label{fig:relprob}
\end{figure}

Apart from the metrics above, we also generate the empirical distribution of relative errors $|\frac{Dst - \hat{Dst}}{Dst}|$ for the \emph{GP-AR}, \emph{GP-ARX}, \emph{NM} and \emph{TL} models.

\subsection{Model Comparison}
